---
title: "Basic functionality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_funcitons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In this vignette we explore the basic function `x13_pickmdl` function of the pickmdl3 pacakge. As its name indicates, this function is a wrapper function that runs the `x13` algorithm from the `rjd3x13` package, but with pickdml rather than automodel as the default model selection procedure. 

We first load a time series to be seasonally adjusted. The series in question is the Norwegian retail index for nace 47.6, \textit{Retail sale of cultural and recreation goods}, between January 2014 and July 2026. We define a specification to be used in the example, namely the default specification 'rsa5c', where we turn off transtive outliers, as this option is rarerly used at Statistics Norway. 

Just as in the `x13` function,  `x13_pickmdl` takes the time series and the specification as inputs. Output is a list with fitted model results, where the most important information is shown in the console. We see that the model of order $(2,1,0)(0,1,1)_s$ has been selected. 

```{r load_data, warning = FALSE, message = FALSE}

library(rjd3toolkit)
library(rjd3x13)
library(pickmdl3)

rti_476 <- pickmdl3::pickmdl_data("norwegian_rti")$rti_476

spec_now <- x13_spec("rsa5c") |>
  set_outlier(outliers.type=c("AO","LS")) 

model_476 <- pickmdl3::x13_pickmdl(ts = rti_476,spec=spec_now)

model_476

```

Further information about the model choiche can be retrieved with the function `ok`, which gives the user a list with three objects. The first object `$ok` tells the user whether the selected model fulfills the three pickmdl criteria at the moment of model selection. The second object `ok_final` says wheter the selected model fulfills the three pickmdl criteria at the end of the time series. Not that there is a difference between the `ok`and `ok_final`only if model selection is done based on a shortened series (See the vignette on refreshment policies). The third object `mdl_nr` says which model on the pickmdl list that was selected. We see that the third model was selected, which of course is the model of order $(2,1,0)(0,1,1)_s$ and this model fulfills the selection criteria. It follows that the first and second models on the list were rejected because they did not pass the selection criteria.  

```{r ok}
pickmdl3::ok(model_476)  
```
   
The calculated pickmodel criteria are available if the option `output` is set to "all". Now we can see that the first two models were rejected because they didn't pass the second criterium, which is indepedent residuals.  

```{r crit_tab}
model_476 <- pickmdl3::x13_pickmdl(ts = rti_476,spec=spec_now,output = "all")  
model_476$crit_tab

```
Note that there are only three models on the list. This is because the algorithm by default stops at the first model that passes the criteria. To force calculation of all five models, set the option `fastfirst = FALSE`: 

```{r fastfirst}
model_476 <- pickmdl3::x13_pickmdl(ts = rti_476,spec=spec_now,output = "all",fastfirst = FALSE)
model_476$crit_tab
```

When the option `output = "all"`, the selected model is available in `$sa`. A list of all the calculated models in the pickmdl list are given in `$all`. 

```{r sa_mult, eval = FALSE }
model_476$sa
model_476$sa_mult
```


Finally, the automodel procedure maybe used even within the `x13_pickmdl` function. When `automdl.enabled = TRUE` the automodel procedure is selected, which gives the exact same result as the `x13` function. We see that in this case the automodel procedure has selected a model that is not on the pickmdl list.

```{r}
model_476_auto <- pickmdl3::x13_pickmdl(ts = rti_476, spec = spec_now, automdl.enabled = TRUE)
model_476_auto 
```


