---
title: "What is pickmodel?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(pickmdl3)
```


Pre-adjustment and forecasting with a sARIMA model is of central importance to the x13-metholdogy, as heterogenous calendar effects and effects of outliers need to be taken into account before seasonal components and trend can be calculated with the filter based methods in x13. In Jdemetra and rjd3, the procedure for selecting an sARIMA model is the automodell procedure. The pickmdl approach differs from this automodel approach by restricting the model choiche to an ordered list of five robust models. The pickmdl selection procedure selects the first model on the list that fulfills three pre-defined criteria. 

The five sARIMA models on the pickmdl list are: 
$$
\begin{aligned}
(0,1,1)(0,1,1)_s  \\
(0,1,2)(0,1,1)_s \\
(2,1,0)(0,1,1)_s \\
(0,2,2)(0,1,1)_s \\
(2,1,2)(0,1,1)_s
\end{aligned}
$$
The tree pre-defined criteria are: 

1. The absolute average percentage error of the extrapolated values within the last three years of data is
less than 15 percent
2. The p-value associated with the fitted model’s Ljung-Box Q-statistic test of the lack of
correlation in the model’s residuals must be greater than 5 percent
3. There are no signs of overdifferencing. There is an indication of overdifferencing if the sum of the non-seasonal MA parameter estimates (for models
with at least one non-seasonal difference) is greater than 0.9.


```{r, include = FALSE}

ts_now <- pickmdl3::pickmdl_data("norwegian_rti")$rti_476

ts_2023 <- window(ts_now,start=start(ts_now),end=c(2023,12))
ts_2024 <- window(ts_now,start=start(ts_now),end=c(2024,12))
ts_jan2025 <- window(ts_now,start=start(ts_now),end=c(2025,1))

spec_now <- x13_spec("rsa5c") |>
  set_transform(fun="Log")|>
  set_outlier(outliers.type=c("AO","LS")) 

oi <- x13(ts_2023,spec=spec_now)


spec_now_dec <- x13_refresh(oi$result_spec,refspec=spec_now,policy="Outliers",period=12,start=c(2024,1))
oi_now_dec <- x13(ts_2024,spec_now_dec)
  
oi2 <- x13(ts_2024,spec=spec_now)
spec_now_dec24 <- x13_refresh(oi2$result_spec,refspec=spec_now,policy="Outliers",period=12,start=c(2025,1))
oi3 <- x13(ts_jan2025,spec_now_dec24)


pick_oi <- pickmdl3::x13_pickmdl(ts_2024,spec=spec_now,identification_end = c(2023,12))
pick_oi2 <- pickmdl3::x13_pickmdl(ts_jan2025,spec=spec_now,identification_end = c(2024,12))
pick_oi3 <- pickmdl3::x13_pickmdl(ts_now,spec=spec_now,identification_end = c(2024,12))


```


# Why use pickmdl? 

By restricting the model choiche to an ordered list of five models, the pickmdl approach prioritizes model stability. The pickmdl choiche may not be the optimal  model, but one or more models on the list will often be of acceptable quality. This may lead to less future revisions of seasonally adjusted data, as the pickmdl approach tends to select the same model if the model in question fits the selection criteria. The search for the model with the optimal fit may lead to model change even though there is only a small difference in quality, with unnecessary revisions as a consequence. 

Let us illustrate this point with an example. We use the Norwegian retail index for nace 47.6  Retail sale of cultural and recreation goods. In accordance with the best practice defined in ESS Guideluines on seasonal adjustment, model identification should be done once a year and the sARIMA model should be kept frozen n line with the chosen refreshment policy . Let us say that model identification is done in January each year, based on data until the preceeding December. Through the year, the model is kept the same in accordance with the Outliers refreshment policy. Thus, each year there is a risk of major revisions, as the sARIMA model may change. 

We see an example of this when we compare the seasonnaly adjusted series in December 2024 and January 2025. The first case is based on the model resulting from the model choiche done in january 2024. The automodel rejected at that point the AIRLINE model and selected the sARIMA model (0,0,1)(0,1,1) as the optimal choiche. In january 2025, a new model identification is done, an at this point the AIRLINE model was still rejected, but the identified model has now changed to (1,0,0)(0,1,1). In figure 1, we compare the resulting seasonally adjusted series. We see that the model change introduces revisions . 
```{r, echo=FALSE}

plot(oi_now_dec$result$final$d11final,xlim=c(2022,2025), ylim=c(80,110),ylab="",xlab="")
lines(oi3$result$final$d11final,col="tomato")
#lines(oi4$result$final$d11final,col="cornflowerblue")

```

With the pickmdl procedure on the other hand, the AIRLINE model is still rejected, but the third model passes the criteria and is selected both in 2024 and 2025, leading to no model change from december to january. As we can see the revisions are now much smaller and pertaining mostly to the end of the series, which is natural. 
```{r, echo=FALSE}



plot(pick_oi$result$final$d11final,xlim=c(2022,2025),ylim=c(80,110),ylab="",xlab="")
lines(pick_oi2$result$final$d11final,col="tomato")
#lines(pick_oi4$result$final$d11final,col="cornflowerblue")

```


A comparison of the revisions is given in figure 3. 


```{r, echo=FALSE}


plot(oi3$result$final$d11final-oi_now_dec$result$final$d11final,xlim=c(2022,2025),ylab="",xlab="") 
lines(pick_oi2$result$final$d11final - pick_oi$result$final$d11final,col="tomato")
abline(h=0,lty=2)


```


