---
title: "Partial concurrent adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{refresh_and_freeze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In accordance with the best practice defined in ESS Guideluines on seasonal adjustment, model identification should be done once a year and the sARIMA model order should be kept in line with the chosen refreshment policy throughout the year. With the pickmdl package, it is easy to implement this best practice in a production process. A date for model selection can be set in the `x13_pickdml`function by the options `identification_end` or `identification_estimate.to`. The `x13_pickmdl` function will then do the model selection based on data only up to and including this date. 

For the sake of illustration, let us say that model selection is to be done each January, based on data until the preceding December, as often is the case at Statistics Norway. We use the norwegian rti in nace 47.6 and the standard specification rsa5c, with outliers turned off to illustrate the option. 


```{r, message=FALSE, warning = FALSE }

library(pickmdl3)

rti_476 <- pickmdl3::pickmdl_data("norwegian_rti")$rti_476

spec_now <- x13_spec("rsa5c") |>
  set_outlier(outliers.type= NULL) 

last_year <- stats::end(rti_476)[1] - 1 

model_now <- x13_pickmdl(rti_476,spec=spec_now,identification_end = c(last_year,12),output="all")

# or equivalently 

model_now <- x13_pickmdl(rti_476,spec=spec_now,identification_estimate.to = paste0(last_year,"-12-01")) 

ok(model_now)

```

Based on the model identification set at the preceding December, the second model on the pickmdl list is chosen by the pickmdl procedure. Compare this to what would have happened if model selection was done with all available data up toJuly 2026. In this case the chosen model is the fifth  model, of order $(2,1,2)(0,1,1)_s$. Model change is thus prevented when we specify identification date for model selection. 

```{r}
model_nowb <- x13_pickmdl(rti_476,spec=spec_now)
```

Notice that there now is a difference between `$ok` and `ok_final` in the output above. Now `$ok` tells us that the selected model satisfied the pickmdl criteria at the identification date, i.e. December 2025. The `ok_final` gives information about  whether the selected model still passes the criteria when the additional data is taken into account. That is, whether the model selected based on data until December 2025 still fits the data in July 2026. We see that that is no longer the case, which is an indication that model change should be expected when the model is to be re-identified in January 2027. 

## Outliers 

In the example above, we didn't consider outliers. The default option in `x13_pickmdl` is that outliers are identified by the shortened series in the cases when model selection is set to a past date. The outliers identified by the shortened series are kept throughout the year, and only new ouliers after the identification date are searched for throughout the year. 

In the example below, we manipulate the series to identify an outlier in June 2026, that is after model selection is done based on data up to and including December 2025.

```{r}

spec_now <- x13_spec("rsa5c") |>
   set_outlier(outliers.type= c("AO","LS"))
  

last_year <- stats::end(rti_476)[1] - 1 

rti_476_fix <- rti_476
rti_476_fix[length(rti_476)-1] <- 10

model_now <- x13_pickmdl(rti_476_fix,spec=spec_now,identification_end = c(last_year,12))

model_now




```

We see that seven outliers are identified in total. Six of these have a date before the model selection date in December 2025 and was therefore identfied in the model selection step. If we take a closer look at the estimation specification for the model in June 2026, we see that the six outliers have been defined as pre-specfied outliers and that the detection span for new outliers  is only from January 2026:

```{r}
model_now$estimation_spec

```
It is possible to let the outlier span be the whole series, even though the model order is identified at a specific date. This is done by setting the option `identify_outliers = FALSE`. Comparing outliers the outliers with the model above, we see that they are somewhat different, as they now all were identified based on the whole available time series. 

```{r}

rar <- x13_pickmdl(rti_476_fix,spec=spec_now,identification_end = c(last_year,12),
                   identify_outliers = FALSE)


```



## Refreshment policy 



The specification at the identification time is the reference specification for the following year. Estimation specification is refreshed according to refreshment policy, as defined in x13_refresh(). Default refreshment policy in the pickmdl package is `Outliers`, where the model order is set, but parameters estimated anew. (Obs: only Outliers available pro tempora) To be extended. 










